LLM Output Compliance with Handcrafted Linguistic Features --- An Experiment

### 2024-11-14 article/main.tex
refer to HLF results in a single way

\textbf{Text style transfer (TST)} is defined as the ``task of transforming the
stylistic manner in which a sentence is written, while preserving the meaning of
the original sentence''~\cite{tst-review-2021}.
This definition can be extended to entire articles or corpuses of text because
these are also the object of linguistic style and
stylistics~\cite{lugea2023stylistics}.
The task of transferring text style has a certain maturity.
The interest in this task is renewed by the advancements made with LLMs.

@@

Our approach stands out through its simplicity and focus.
We propose engineering instruction prompts for LLMs, a straightforward strategy
distinct from complex alternatives.
With discussions on style transfer or author verification beyond this paper's
scope, we present a method controlling LLM text generation through the use of
HLFs.

@@

Our experiment analyses LLM output when asked to generate new text based on
given input text and instructions derived from HLFs.
We compute statistics of HLFs on a corpus to derive the text generation
instructions.
To run the experiment we must select LLMs to evaluate and find datasets from
which we can derive corpuses.
Then we must also select the HLFs for our experiment.
Before going over our selection for the experiment's parameters, let us go
through the experimental process.

@@

The second scenario consists in generating text based on input text, but with
the aid of some examples from the corpus.
The examples are appended to the instruction prompt used in the first scenario.
The number of examples is chosen so that we do not overrun the context window of
the LLM.\@
Some context windows are smaller than others, settling the number of examples at
three.
Adding examples to the LLM's instruction prompt also makes the provenance of the
input text more important.
Therefore we experiment with two variants of the second scenario: one with input
text extraneous to the corpus and one with input text from the corpus.
The text from outside the corpus is the same text we used in the first scenario.
The text from within the corpus is chosen randomly once at the start of the
experiment.
Following the process from the first scenario results in obtaining baseline HLF
values for each variant of the second scenario.

@@

In a similar manner to how we generated the baselines, we run 10 trials for each
scenario and variant.
We compare each baseline with the results obtained by using the corresponding
HLF prompt.
From here on we refer to the results obtained by using prompts with HLF
instructions as `HLF-enhanced results', `HLF results' or even simply `results'.

@@

\begin{table*}[ht!]
    \setlength\tabcolsep{6pt}
    \centering
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{@{}llllll@{}}
            \toprule
            LFTK ID         & Name                              & Family           & Domain           & Abstraction     & Style Level\footnotemark\\ \toprule
            t\_word         & total words                       & wordsent         & surface          & very concrete   & lexical                 \\
            t\_sent         & total sentences                   & wordsent         & surface          & very concrete   & syntactic               \\ \bottomrule
            n\_uverb        & total number of unique adjectives & partofspeech     & syntax           & concrete        & semantic                \\
            n\_uadj         & total number of unique adjectives & partofspeech     & syntax           & concrete        & syntactic               \\ \bottomrule
            simp\_ttr       & simple type token ratio           & typetokenratio   & lexico-semantics & regular         & pragmatic               \\
            a\_verb\_pw     & number of verbs per word          & partofspeech     & surface          & regular         & syntactic               \\ \bottomrule
            corr\_adj\_var  & corrected adjective variation     & lexicalvariation & lexico-semantics & abstract        & lexical                 \\
            corr\_verb\_var & corrected verb variation          & lexicalvariation & lexico-semantics & abstract        & lexical                 \\ \bottomrule
            fkgl            & Flesch-Kincaid grade level        & readformula      & surface          & highly abstract & pragmatic               \\
            a\_kup\_pw      & Kuperman age of aquisition        & worddiff         & lexico-semantics & highly abstract & lexical                 \\ \bottomrule
        \end{tabular}%
    }
    \caption{Handcrafted Linguistic Features}\label{table-hlf}
\end{table*}
\footnotetext{as described in the figure in~\cite{lugea2023stylistics}, chapter 1, page 7}

@@

CNN/DailyMail contains stories and news with diverse styles.
The style varies from one author to another and it also follows the publisher's
editorial style.
The articles are more formal, longer and more complex in structure than Yelp
reviews which makes this dataset a good alternative.
The corpus that is based on this dataset contains the first 1000 CNN articles
from the test split of the dataset.

@@

Determining whether the baseline results differ from the HLF-enhanced results is
a classic use case for the Kolmogorov-Smirnov
test~\cite{kolmogorov1933,smirnov1939}, as implemented in the
Scipy library~\cite{2020SciPy-NMeth}.
We consider that the results are different for a resulting $p\_value \le 0.05$.

@@

The results on the CNN article corpus showcased in Table~\ref{table-prompt-1-cnn-dailymail}
shows that LLMs take into account instructions concerning the HLFs of their
output.

The results on the review corpus generated from the Yelp dataset are similarly
significant, but the HLF instructions determine outcomes that are further away
from the target.

@@

There are two variants of this scenario: input text from within or from outside
the corpus.
Table~\ref{table-prompt-2-cnn-dailymail} contains results for the first variant:
input from the corpus.
We notice these results are similar to those in the first experiment scenario.

@@

A possible explanation for having fewer significant differences in results might
lie in the closeness to the target.
As we can see in Figure~\ref{fig-p2-gpt-twords}, this is not always the case.
In~\cref{fig-p2-command-r-simpttr,fig-p2-gemini-a-kup-pw} we observe regressions
in performance with significant differences between baseline and result.
Even so, the results obtained in our first scenario are largely confirmed in the
first variant of our second scenario.

\begin{figure*}[ht!]
    \centering
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2/prompt_2-gpt-cnn_dailymail/prompt_2-gpt-cnn_dailymail_t_word.png}
        \caption{GPT\\Total Words}\label{fig-p2-gpt-twords}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2/prompt_2-command_r-cnn_dailymail/prompt_2-command_r-cnn_dailymail_simp_ttr.png}
        \caption{Command-R+\\Simple TTR}\label{fig-p2-command-r-simpttr}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2/prompt_2-gemini-cnn_dailymail/prompt_2-gemini-cnn_dailymail_a_kup_pw.png}
        \caption[center]{Gemini\\Kuperman Age of Aquisition}\label{fig-p2-gemini-a-kup-pw}
    \end{minipage}
\end{figure*}

@@

\begin{figure*}[ht!]
    \centering
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2_ifd/prompt_2-claude3-cnn_dailymail/prompt_2-claude3-cnn_dailymail_n_uadj.png}
        \caption{Claude3\\Number of Unique Adjectives}\label{fig-p2-ifd-claude3-nuadj}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2_ifd/prompt_2-gemini-cnn_dailymail/prompt_2-gemini-cnn_dailymail_t_word.png}
        \caption{Gemini\\Total Words}\label{fig-p2-ifd-gemini-twords}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2_ifd/prompt_2-llama3_70b-cnn_dailymail/prompt_2-llama3_70b-cnn_dailymail_a_kup_pw.png}
        \caption[center]{Llama3\\Kuperman Age of Aquisition}\label{fig-p2-ifd-llama3-a-kup-pw}
    \end{minipage}
\end{figure*}

@@

In terms of our first question --- whether HLF instructions impact an LLM's
writing, the response is affirmative.
Not all LLMs are equally susceptible to HLF instructions, though.
The experiment design involves choices that rely on the presence of a corpus and
the usage of a fixed external input text.
This is relevant as evidentiated by the difference in results obtained using the
Yelp reviews corpus when compared to the results obtained on the CNN article
corpus.
While similarly significant, the HLF results on Yelp reviews are mostly worse
than their corresponding baselines failing to confirm the positive impact of HLF
instructions.
The importance of the choice of the external text is highlighted by the better
results obtained using an input text from the corpus.

This leads us to our second question: how much do HLF instructions impact an
LLM's writing?
The metrics we compute dismiss the idea that HLF instructions have a consistent
desired impact on the generated text.
There isn't conclusive evidence that improving the LLM's odds of producing a
better baseline results in a smaller difference between baseline and result
relative to the target.
More importantly, there is evidence that HLF instructions have an adverse effect
when comparing the results on the Yelp reviews dataset.

Highly abstract HLF instructions are unlikely to have the desired effect.
Furthermore, the impact tends to be adverse.
This suggests that LLMs are more adept at following instructions regarding HLFs
when those instructions are more concrete.

The choice of input text in relation to the target HLF values also seems
important.
LLMs don't yet seem able to cover the gulf between Oscar Wilde poems and the
average Yelp review in terms of linguistic features.
We surmise that even if LLMs are able to generate text that complies to certain
target HLFs, there is a limit to this ability.

Finally, selecting HLFs that are in tension with one another will affect the
LLMs ability to comply.
This is highlighted in the experiment with age of acquisition versus adjective
variation.

>>>>

\textbf{Text style transfer (TST)} is defined as the ``task of transforming the
stylistic manner in which a sentence is written, while preserving the meaning of
the original sentence''~\cite{tst-review-2021}.
This definition can be extended to entire articles or corpora of text because
these are also the object of linguistic style and
stylistics~\cite{lugea2023stylistics}.
The task of transferring text style has a certain maturity.
The interest in this task is renewed by the advancements made with LLMs.

@@

Our approach stands out through its simplicity and focus.
We propose engineering instruction prompts for LLMs, a straightforward strategy
distinct from complex alternatives.
With discussions on style transfer or author verification beyond this paper's
scope, we present a method for controlling LLM text generation through the use
of HLFs.

@@

Our experiment analyses LLM output when asked to generate new text based on
given input text and instructions derived from HLFs.
We compute statistics of HLFs on a corpus to derive the text generation
instructions.
To run the experiment we must select LLMs to evaluate and find datasets from
which we can derive corpora.
Then we must also select the HLFs for our experiment.
Before going over our selection for the experiment's parameters, let us go
through the experimental process.

@@

The second scenario consists in generating text based on input text, but with
the aid of some examples from the corpus.
The examples are appended to the instruction prompt used in the first scenario.
The number of examples is chosen so that we do not overrun the context window of
the LLM.\@
Some context windows are smaller than others, setting the number of examples at
three.
Adding examples to the LLM's instruction prompt also makes the provenance of the
input text more important.
Therefore we experiment with two variants of the second scenario: one with input
text extraneous to the corpus and one with input text from the corpus.
The text from outside the corpus is the same text we used in the first scenario.
The text from within the corpus is chosen randomly once at the start of the
experiment.
We repeat the same process from the first scenario to obtain baseline HLF values
for each variant of the second scenario.

@@

In a similar manner to how we generated the baselines, we run 10 trials for each
scenario and variant.
We compare each baseline with the results obtained by using the corresponding
HLF prompt.
From here on we refer to the results obtained by using prompts with HLF
instructions as `HLF results'.

@@

\begin{table*}[ht!]
    \setlength\tabcolsep{6pt}
    \centering
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{@{}llllll@{}}
            \toprule
            LFTK ID         & Name                              & Family           & Domain           & Abstraction     & Style Level\footnotemark\\ \toprule
            t\_word         & total words                       & wordsent         & surface          & very concrete   & lexical                 \\
            t\_sent         & total sentences                   & wordsent         & surface          & very concrete   & syntactic               \\ \bottomrule
            n\_uverb        & total number of unique verbs      & partofspeech     & syntax           & concrete        & syntactic               \\
            n\_uadj         & total number of unique adjectives & partofspeech     & syntax           & concrete        & syntactic               \\ \bottomrule
            simp\_ttr       & simple type token ratio           & typetokenratio   & lexico-semantics & regular         & pragmatic               \\
            a\_verb\_pw     & number of verbs per word          & partofspeech     & surface          & regular         & syntactic               \\ \bottomrule
            corr\_adj\_var  & corrected adjective variation     & lexicalvariation & lexico-semantics & abstract        & lexical                 \\
            corr\_verb\_var & corrected verb variation          & lexicalvariation & lexico-semantics & abstract        & lexical                 \\ \bottomrule
            fkgl            & Flesch-Kincaid grade level        & readformula      & surface          & highly abstract & pragmatic               \\
            a\_kup\_pw      & Kuperman age of acquisition       & worddiff         & lexico-semantics & highly abstract & lexical                 \\ \bottomrule
        \end{tabular}%
    }
    \caption{Handcrafted Linguistic Features}\label{table-hlf}
\end{table*}
\footnotetext{as described in the figure in~\cite{lugea2023stylistics}, chapter 1, page 7}

@@

CNN/DailyMail contains stories and news with diverse styles.
The style follows the publisher's guidelines, but, within those guidelines, it
varies for each author.
The articles are more formal, longer and more complex in structure than Yelp
reviews which makes this dataset a good alternative.
The corpus that is based on this dataset contains the first 1000 CNN articles
from the test split of the dataset.

@@

We can determine whether the baseline results differ from the HLF results using
the Kolmogorov-Smirnov test~\cite{kolmogorov1933,smirnov1939}, as implemented in
the Scipy library~\cite{2020SciPy-NMeth}.
We consider that the results are different for a $p\_value \le 0.05$.

@@

Table~\ref{table-prompt-1-cnn-dailymail} shows that LLMs take into account
HLF instructions when running the experiment.
On the CNN article corpus, the HLF instructions lead to HLF results that are
closer than the baseline to the target HLF.\@
On the Yelp dataset, the HLF instructions generate results that are further away
than the baseline from the target HLF.\@

@@

There are two variants of this scenario depending on the provenance of the input
text.
The input text may be from outside the corpus or a text from within the corpus.
Table~\ref{table-prompt-2-cnn-dailymail} contains results for the first variant:
input from outside the corpus.
We notice the HLF results tend to be closer to the target than the baseline
results.
This picture is similar to the one we saw in the first scenario.

@@

On the other hand, there are fewer significant differences between baseline and
HLF results.
A possible explanation for this situation might lie in the closeness of the
input prompt (which contains examples from the corpus)to the target.
As we can see in Figure~\ref{fig-p2-gpt-twords}, this is not always the case.
In~\cref{fig-p2-command-r-simpttr,fig-p2-gemini-a-kup-pw} we observe regressions
in performance with significant differences between baseline and result.
Even so, the results obtained in our first scenario are largely confirmed in the
first variant of our second scenario.

\begin{figure*}[ht!]
    \centering
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2/prompt_2-gpt-cnn_dailymail/prompt_2-gpt-cnn_dailymail_t_word.png}
        \caption{GPT\\Total Words}\label{fig-p2-gpt-twords}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2/prompt_2-command_r-cnn_dailymail/prompt_2-command_r-cnn_dailymail_simp_ttr.png}
        \caption{Command-R+\\Simple TTR}\label{fig-p2-command-r-simpttr}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2/prompt_2-gemini-cnn_dailymail/prompt_2-gemini-cnn_dailymail_a_kup_pw.png}
        \caption[center]{Gemini\\Kuperman Age of Acquisition}\label{fig-p2-gemini-a-kup-pw}
    \end{minipage}
\end{figure*}

@@

\begin{figure*}[ht!]
    \centering
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2_ifd/prompt_2-claude3-cnn_dailymail/prompt_2-claude3-cnn_dailymail_n_uadj.png}
        \caption{Claude3\\Number of Unique Adjectives}\label{fig-p2-ifd-claude3-nuadj}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2_ifd/prompt_2-gemini-cnn_dailymail/prompt_2-gemini-cnn_dailymail_t_word.png}
        \caption{Gemini\\Total Words}\label{fig-p2-ifd-gemini-twords}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2_ifd/prompt_2-llama3_70b-cnn_dailymail/prompt_2-llama3_70b-cnn_dailymail_a_kup_pw.png}
        \caption[center]{Llama3\\Kuperman Age of Acquisition}\label{fig-p2-ifd-llama3-a-kup-pw}
    \end{minipage}
\end{figure*}

@@

In terms of our first question --- whether HLF instructions impact an LLM's
writing, the response is affirmative.
Not all LLMs are equally susceptible to HLF instructions, though.
The experiment design involves choices that rely on the presence of a corpus and
the usage of a fixed external input text.
This is relevant as evidentiated by the difference in HLF results obtained using
the Yelp reviews corpus when compared to the HLF results obtained on the CNN
article corpus.
While similarly significant, the HLF results on Yelp reviews are mostly worse
than their corresponding baselines failing to confirm the positive impact of HLF
instructions.
The importance of the choice of the external text is highlighted by the better
HLF results obtained using an input text from the corpus.

Let us move on to how much HLF instructions impact an LLM's writing.
The metrics we compute dismiss the idea that HLF instructions have a consistent
desired impact on the generated text.
There isn't conclusive evidence that if we improve the LLM's chances of
producing a better baseline, this will result in a smaller difference between
baseline and HLF result relative to the target HLF.
In fact, there is evidence that HLF instructions have an adverse effect when
examining the HLF results obtained on the Yelp reviews dataset.
This is especially true of highly abstract HLF instructions.
LLMs do not seem to follow these instructions well in most scenarios.
The impact of highly abstract HLF instructions tends to be adverse.

The choice of input text in relation to the target HLF values is consequential.
LLMs don't yet seem able to cover the gulf between Oscar Wilde poems and the
average Yelp review in terms of linguistic features.
We surmise that even if LLMs are able to generate text that complies to certain
target HLFs, there is a limit to this ability.
